EuPPollNet: A European database of plant-pollinator networks

### 2024-05-27 Manuscript/EuPPollNet_Ms.qmd
edits

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at European level by combining the recently published checklists of bees [@ghisbain2023], syrphids [@kovcic2023] and butterflies [@wiemers2018]; (ii) then, we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we programmatically fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists.

@@

EuPPollNet offers the largest set of plant-pollinator networks and plant-pollinator studies compiled to date at European level. The database contains `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species with over a million of interaction records. While the overall sampling coverage of species and interactions is relatively high across the sampled sites, the taxonomic coverage of plants and the main pollinator groups at the European level is still relatively low (e.g., `r round(plant_coverage, 1)`% for flowering plants and `r round(bee_coverage, 1)`% for bee species). Given that most of the plant-pollinator networks from the database are sampled on pastures, and habitat heterogeneity is a crucial factor in understanding pollinator diversity at European level [@kleijn2015; @hass2018; @martinez2022], adding studies on other habitat types is likely to result in a rapid increase of the coverage of plant and pollinator species and their interactions. Plant and pollinator species were rarely shared across multiple sites, indicating that there are few "common" species and many "rare" ones at the metaweb or continental level. This high number of "rare" species results in an upward slope of the species or interaction accumulation curves [@thompson2003]. In other words, minimal sampling efforts are capturing a substantial number of species and interactions, but achieving a comprehensive inventory will require of numerous sampling events within and across habitats, particularly for plant-pollinator interactions.

Bees conducted the majority of the interactions at the metaweb level, but their relative relevance changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were especially fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only species rich in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted in average a third of the total interactions per network. This proportion is higher that the one found when considering only natural communities at a global scale [~13%\; @hung2018], which highlights the need of accounting for the potential negative impact of honey bees to wild pollinator communities when managing pollinator communities in Europe [@steffan2000; @herrera2020].

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species almost doubled the one of plants in the database. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centred sampling is likely to increase the plant:pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. As expected [@jordano1987], we found that network connectance decreased exponentially with the number of species. Overall, latitude did not explain major changes in corrected connectance or normalised nestedness. However, consistent with @olesen2002, corrected connectance was lower at higher latitudes, while normalised nestedness, which is negatively correlated with corrected connectance, increased towards higher latitudes. Thus, the potential ecological explanations that can help understanding these associations can be similar. For instance, networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019]. Further, the level of pollinator generalization seems to be higher at lower latitudes [@schleuning2012]. Therefore, these different factors can increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected networks but less nested ones in comparison to networks at higher latitudes. In addition, while one third of networks were more nested than expected by chance, two thirds of them did not show statistical difference with null expectations. This goes against the common view that most plant-pollinator networks are nested [@bascompte2003], and supports the idea that despite a strong tendency towards it, being nested is not the rule fo this large set of plant-pollinator networks.

Despite this database covering a wide range of habitats across `r country_number` countries, it contains geographical biases that can impact our understanding of plant-pollinator communities [@hughes2021]. For instance, most plant-pollinator networks belongs to central Europe, while Eastern Europe and the Mediterranean region are under-represented. This is consistent with previous studies which also report lack of plant-pollinator data for those regions [REFs], highlighting that this database reflects existing patterns in data availability despite the absence of a systematic search for studies. This is especially relevant as Eastern Europe is expected to continue developing and use their vast natural resources [@bennett2018], and the Mediterranean region is likely to be severely impacted by climate change [@jaworski2022; @duchenne2020; @pareja2023]. These areas are well known for their rich pollinator diversity [REFs], and their under-representation is likely contributing to the low taxonomic coverage of this database at the European level.

@@

Although some of the most well studied countries in Europe (e.g., Belgium or The Netherlands) seem to have suffered most of the plant and pollinator biodiversity loss at the end of the 20th century [@carvalheiro2013], plant-polllinator communities in Europe and across the globe still suffer from major threads such as climate change [REFs], land use change [REFs], or the introduction of alien species [@vila2009; @vanbergen2018].

@@

>>>>

For pollinators: (i) we first created a checklist of species names for the most representative pollinator groups at the European level by combining the most up to date published checklists of bees and syrphids [@reverte2023] with the one of butterflies [@wiemers2018]; (ii) then, we compared pollinator species names against the checklist and recovered some unmatched cases with restrictive fuzzy matching by using **stringdist** package [@van2014]; (iii) we programmatically fixed unmatched records when necessary and retrieved the taxonomic information for all species from GBIF; (iv) we fixed the non-found cases in the GBIF checklist and made sure that all species names from bees, syrphids and butterflies were named according to their respective species checklists.

@@

EuPPollNet offers the largest set of plant-pollinator networks and plant-pollinator studies compiled to date at European level. The database contains `r prettyNum(plant_spp_number, big.mark=",")` plant and `r prettyNum(pollinator_spp_number, big.mark=",")` pollinator species with over a million of interaction records. While the overall sampling coverage of species and interactions is relatively high across the sampled sites, the taxonomic coverage of plants and the main pollinator groups at the European level is still relatively low (i.e., `r round(plant_coverage, 1)`% for flowering plants and `r round(bee_coverage, 1)`% for bee species). Given that most of the plant-pollinator networks from the database are sampled on pastures, and habitat heterogeneity is a crucial factor in understanding pollinator diversity at European level [@kleijn2015; @hass2018; @martinez2022], adding studies on other habitat types is likely to result in a rapid increase of the coverage of plant and pollinator species and their interactions. However, plant and pollinator species were rarely shared across multiple sites, indicating that there are few "common" species and many "rare" ones at the metaweb or continental level. This high number of "rare" species results in an upward slope of the species or interaction accumulation curves [@thompson2003]. In other words, minimal sampling efforts are capturing a substantial number of species and interactions, but achieving a comprehensive inventory will require of numerous sampling events within and across habitats, particularly for plant-pollinator interactions.

Bees conducted the majority of the interactions at the metaweb level, but their relative relevance changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were especially fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only species rich in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted in average a third of the total interactions per network. Indeed, this proportion is higher that the one found when considering only natural communities at a global scale [~13%\; @hung2018], which highlights their key role as potential pollinator of many flowering plants. However, high honey bee densities can have negative impacts on wild pollinators and the plants they visit [@steffan2000; @geldmann2018; @herrera2020], emphasising the need to evaluate these impacts and implement actions if necessary.

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species almost doubled the one of plants in the database. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centred sampling is likely to increase the plant:pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. As expected [@jordano1987], we found that network connectance decreased exponentially with the number of species. Overall, latitude did not explain major changes in residual connectance or normalised nestedness across networks. However, consistent with @olesen2002, residual connectance -the deviation from the expected connectance for a given network size- was lower at higher latitudes, while normalised nestedness, which is negatively correlated with residual connectance, increased towards higher latitudes. Thus, giving their association and that both metrics capture structural properties of species interactions, the potential ecological explanations that can help understanding their latitudinal trends can be similar. For instance, networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019], and the overall level of pollinator generalization seems to be higher at lower latitudes [@schleuning2012]. Therefore, we hypothesise that these different factors can increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected networks but less nested ones in comparison to networks at higher latitudes. This higher proportion of generalists could result in a more even number of links per species, deviating more from a perfect nested or hierarchical structure of the network. Finally, while one third of networks were more nested than expected by chance, two thirds did not show statistical difference with null expectations. Therefore, while this supports the idea that plant-pollinator networks tend to be nested [@bascompte2003], this result suggests that nestedness could be a less prevalent feature than previously thought for plant-pollinator networks.

Despite this database covering a wide range of habitats across `r country_number` countries, it contains geographical biases that can impact our understanding of plant-pollinator communities [@hughes2021]. For instance, most plant-pollinator networks belongs to central Europe, while Eastern Europe and the Mediterranean region are under-represented. This is consistent with previous studies which also report lack of plant-pollinator data for those regions [@bennett2018; @marshall2024], highlighting that this database shows existing patterns in data availability despite the absence of a systematic search for studies. This is especially relevant as Eastern Europe is expected to continue developing and use their vast natural resources [@sutcliffe2015], and the Mediterranean region is likely to be severely impacted by climate change [@jaworski2022; @duchenne2020; @pareja2023]. These areas are well known for their rich pollinator diversity [@milivcic2018; @reverte2023], and their under-representation is likely contributing to the low taxonomic coverage of this database at the European level. Although some of the most well studied countries in Europe (e.g., Belgium or The Netherlands) seem to have suffered most of the plant and pollinator biodiversity loss at the end of the 20th century [@carvalheiro2013], plant-polllinator communities in Europe and across the globe still suffer from major threads such as climate change [REFs], land use change [@reidsma2006; @batary2015], or the introduction of alien species [@vila2009; @vanbergen2018]. Therefore, continuous monitoring programs of plant-pollinator communities are needed in order to evaluate

are essential to understand temporal and spatial changes in order to preserve their biodiversity.

@@

@@

Hence, this emphasize how this database can help understanding ecological questions from plant-pollinator networks and its potential implications for future research questions.
